% $Id: bom.tex 9493 2021-10-26 22:07:41Z mskala $

%
% MSK 010 bill of materials
% Copyright (C) 2017, 2021  Matthew Skala
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, version 3.
%
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
%
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
% Matthew Skala
% https://northcoastsynthesis.com/
% mskala@northcoastsynthesis.com
%

\onecolumn
\chapter{Bill of materials}

{\centering
\fbox{This table is not a substitute for the text instructions.}

\begin{longtable}{rp{1in}cp{3in}}
  \textbf{Qty} & \textbf{Ref} & \textbf{Value/Part No.} & \\ \hline \endhead
\input{bomdata.tex}
\end{longtable}\par}

Not listed above, but also needed:  one front panel, two circuit boards, two
14-pin DIP sockets for the ICs, solder, and incidental supplies such as
alcohol for cleaning.

Fixed resistors should be 1\%\ metal film throughout.  Film capacitors
should be 5\%\ or better.  These tolerances are important to guarantee
oscillation.  RoHS-certified zinc-plated steel hardware is recommended, not
stainless steel because of galvanic-corrosion incompatibility with aluminum
parts.

Resistor and capacitor reference designators shown are for the ``A'' variant
of the module; install them in a different arrangement, according to the
instructions in the relevant chapters, for other variants.

North Coast Synthesis Ltd.\ kits or assembled modules sometimes contain
other parts with equivalent or better specifications rather than exact
manufacturers and part numbers shown here.  \twocolumn
